Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-20 article.tex
some modifs : version ready for Damien + new images

\begin{abstract}
In this work, we propose a simple and mathematically motivated method for the
problem of connectome inference in calcium imaging data. The proposed algorithm
consists in i) processing the raw signals to detect neural peak activities and ii)
inferring the degree of association between neurons from partial correlation
statistics.  This paper summarizes the methodology that led us to win the
Connectomics Challenge, proposes a simplified version of our method and finally
discusses our results with respect to other inference methods.
\end{abstract}

@@

\textcolor{red}{To improve (style):} The human brain is a complex biological organ made of 100 billions of neurons,
each connected to 7000 other neurons on average. Unfortunately, direct
observations of the connectome, the wiring diagram of the brain, is not yet
technically feasible. Without being perfect, calcium imaging currently allows
the real-time and simultaneous observation of neuron activities from thousands
of neurons, producing individual time series representing their fluorescence
intensity. From these data, the connectome inference problem consists in
retrieving the synaptic connections between neurons on the basis of the
fluorescence time series. In particular, this problem is often made difficult
because of experimental issues, including masking effects (making some of the
neurons not to be observed or confounded with others), the low sampling rate of
the optical device with respect to the neural activity speed or the slow decay
of fluorescence.

@@

\begin{figure}
\centering
\includegraphics[width=0.4\linewidth]{images/fig_filtering}
\caption{Signal processing process for extracting peaks from the raw fluorescence data.}
\label{fig:filtered-signal}
\end{figure}

@@

At this step, the processed signal only contains clean spikes. However, when a
large part of the network is firing, detecting pairwise neuron interactions is a
difficult task. \textcolor{red}{G: say that neurons may fire because of global
high activity, without any actual cause -- see generator.}
Accordingly, in order to reduce false interaction detection,
peak intensities at time $t$ are regularized by the global activity of all
neurons, thereby reducing the importance of peaks identified during global high activity periods:
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}.
\end{align}

@@

\begin{figure}[bth]
\centering
\subfigure[ROC curves]{\includegraphics[width=0.45\textwidth]{images/curve_roc} \label{fig:roc_curve}}
\subfigure[Precision-recall curves]{\includegraphics[width=0.45\textwidth]{images/curve_pr} \label{fig:pr_curve}}
\caption{The weighted average of partial correlation matrix, winning approach
         of the challenge, yields better than AUROC and AUPRC than all
         other state-of-the-art methods. (parameters: $\tau=0.11$ except for
         averaged partial correlation method, partial correlation used
         filter $f_1$).}
\label{fig:curves}
\end{figure}

@@

The best method only infers the \textbf{undirected} network by giving the
same connectivity score to both edges $(i,j)$ and $(j,i)$. This is quite
surprising considering the problem statement.

If the undirected network is perfectly inferred, it would already yield
high AUROC score for instance $\approx{}0.996$ on \textit{normal-1}. Trying
to infer edge directivity is likely to be counter-productive or yield
small gain.
By contrast, the same experiment with AUPRC would yield
to a score of $\approx 0.789$ on \textit{normal-1}. Thus inferring
correctly directivity would be more rewarding in AUPRC term.
Interestingly, methods that takes into account directivity, namely GTE and
GENIE3, improve their AUROC and AUPRC by giving the same way to directed edges.
(in particular see GTE and GTE(undirected) in Figure \ref{fig:pr_curve}
\textcolor{red}{AJ: Figure \ref{fig:curves}  contradicts this message,
especially pr-curve. AS:And now?} and Table \ref{tab:tab1} in Appendix \textcolor{red}{REF}).
Trying to detect the link direction is more penalizing than rewarding.

The area under the ROC curve is not consistent metric
with respect to the problem statement. A clear hierarchy is also difficult to
derive based on ROC curve of Figure \ref{fig:curves}. We advice to avoid the AUROC
for the assessment of network inference method. An alternative metric would
be the area under the precision-recall curve (AUPRC). Those
conclusions are in line with \cite{schrynemackers2013protocols}.

>>>>

\begin{abstract} In this work, we propose a simple and mathematically
motivated method for the problem of connectome inference in calcium imaging
data. The proposed algorithm is made of two steps. First, processing the raw
signals to detect neural peak activities. Second, inferring the degree of
association between neurons from partial correlation statistics. This paper
summarizes the methodology that led us to win the Connectomics Challenge,
proposes a simplified version of our method and finally discusses our results
with respect to other inference methods. \end{abstract}

@@

%%%%%%% !!!!!!!! \textcolor{red}{To improve (style):} !!!! %%%%%%%%
The human brain is a complex biological
organ made of 100 billions of neurons, each connected to 7000 other neurons on
average. Unfortunately, direct observations of the connectome, the wiring
diagram of the brain, is not yet technically feasible. Without being perfect,
calcium imaging currently allows the real-time and simultaneous observation of
neuron activities from thousands of neurons, producing individual time series
representing their fluorescence intensity. From these data, the connectome
inference problem consists in retrieving the synaptic connections between
neurons on the basis of the fluorescence time series. In particular, this
problem is often made difficult because of experimental issues, including
masking effects (making some of the neurons not to be observed or confounded
with others), the low sampling rate of the optical device with respect to the
neural activity speed or the slow decay of fluorescence.
%%%%%%%%%%% !!!!!!!!!!!!!! %%%%%%%%%%%%%%%%

@@

\begin{figure}
\centering
\subfigure[Original]{\includegraphics[width=0.3\textwidth]{images/original_curve.pdf} \label{fig:original_curve}}
\subfigure[Low-pass $f_1(x_i^t)$]{\includegraphics[width=0.3\textwidth]{images/filter_curve.pdf} \label{fig:lp_curve}}\\
\subfigure[High-pass $g(x_i^t)$]{\includegraphics[width=0.3\textwidth]{images/diff_curve.pdf} \label{fig:hp_curve}}
\subfigure[Thresholding $h(x_i^t)$]{\includegraphics[width=0.3\textwidth]{images/threshold_curve.pdf} \label{fig:threshold_curve}}\\

\subfigure[Weights $w(x_i^t)$]{\includegraphics[width=0.3\textwidth]{images/weights_curve.pdf} \label{fig:original_curve}}
\caption{Signal processing process for extracting peaks from the raw fluorescence data.}
\label{fig:filtered-signal}
\end{figure}

% \begin{figure}
% \centering
% \includegraphics[width=0.4\linewidth]{images/fig_filtering}
% \caption{Signal processing process for extracting peaks from the raw fluorescence data.}
% \label{fig:filtered-signal}
% \end{figure}

@@

At this step, the processed signal only contains clean spikes. However, when a
large part of the network is firing, detecting pairwise neuron interactions is a
difficult task. Futhermore, it is even possible that neurons fire because of global high activity, without actual cause (see \cite{stetter2012model}).
% \textcolor{red}{G: say that neurons may fire because of global
% high activity, without any actual cause -- see generator.}
Accordingly, in order to reduce false interaction detection,
peak intensities at time $t$ are regularized by the global activity of all
neurons, thereby reducing the importance of peaks identified during global high activity periods:
\begin{align}
 w(x^{t}_i) &= (x^{t}_i + 1 )^{1 + \frac{1}{\sum_{j} x^{t}_j}}.
\end{align}

@@

\begin{figure}[bth]
\centering
\subfigure[ROC curves]{\includegraphics[width=0.45\textwidth]{images/curve_roc} \label{fig:roc_curve}}
\subfigure[Precision-recall curves]{\includegraphics[width=0.45\textwidth]{images/curve_pr} \label{fig:pr_curve}}
\caption{The weighted average of partial correlation matrix, winning approach
         of the challenge, yields better than AUROC and AUPRC than all
         other state-of-the-art methods. (parameters: $\tau=0.11$ except for
         averaged partial correlation method, partial correlation used
         filter $f_1$). }
\label{fig:curves}
\end{figure}

@@

The best method only infers the \textbf{undirected} network by giving the same
degree of association to both edges $(i,j)$ and $(j,i)$. This is quite
surprising considering the problem statement. Trying to direct links could be
more penalizing than rewarding. For example, symmetrizing the degree of
association matrix does not necessarily degrade AUPRC and AUROC.

If the inferred network corresponds to the undirected network, it would
already yield high AUROC score for instance $\approx 0.996$ on
\textit{normal-1}. Infering edge directivity is likely to be counter-
productive or yield small gain. By contrast, the same experiment with AUPRC
would yield to a score of $\approx 0.789$ on \textit{normal-1}. Thus inferring
correctly directivity would be more rewarding in AUPRC term.

% Interestingly, methods that takes into account directivity, namely GTE and
% GENIE3, improve their AUROC and AUPRC by giving the same way to directed edges.
% (in particular see GTE and GTE(undirected) in Figure \ref{fig:pr_curve}
% \textcolor{red}{AJ: Figure \ref{fig:curves}  contradicts this message,
% especially pr-curve. AS:And now?} and Table \ref{tab:tab1} in Appendix \textcolor{red}{REF}).
% Trying to detect the link direction is more penalizing than rewarding.

The area under the ROC curve is not consistent metric
with respect to the problem statement. A clear hierarchy is also difficult to
derive based on ROC curve of Figure \ref{fig:roc_curve}. We advice to avoid the AUROC
for the assessment of network inference method. An alternative metric would
be the area under the precision-recall curve (AUPRC). Those
conclusions are in line with \cite{schrynemackers2013protocols}.
